\mythesistitleNOTSYNPOSIS \tnote[fyi]{Synopsis title is ``Generalization of Consecutive Ones Property of Binary Matrices''}

### 2012-01-18 src/020chap2.tex
first draft of PQR trees done. next is cert algo and icpia.

\begin{definition}%
  \label{def:matrixsetsystem}{\emph{\Cop for set systems}}%
  % [\emph{Set system of a binary matrix}.]
   Let $M$ be a binary matrix of order
  $n \times m$ % , $r_i, i \in [n]$ be the rows in $M$
  and $\{c_i \mid i \in [m]\}$ be the columns in $M$. A set system
  $\cF_{M} = \{S_i \mid S_i \subseteq [n], i \in [m]\}$ is defined
  such that for every column $c_i$ of $M$, set $S_i = \{j \mid m_{ji}
  = \un \}$. The collection $\cF_M$ is the {\em set system of binary
    matrix} $M$.\\
  A set system $\cF$ from universe $U = \{1, 2, \ldots, n\}$ has the
  {\em \cop} if it is possible to assign a linear order, \ie a
  permutation $\sigma = u_1u_2\ldots u_n$ to $U$ \stt each set $S \in
  \cF$ appears as a consecutive subsequence $u_{i}u_{i+1}\ldots
  u_{i+k-1}$ (also called an interval) of $\lambda$ in the linear order, where
  positive integers $k = |S|$ and $i \le n+1-k$.
\end{definition}

@@

Generalized \PQtree or \gPQtree is a data structure defined in \cite{n89} to
represent all orthogonal sets of a set system $\cF$. {\em Orthogonal
  sets} of $\cF$ are sets that do not overlap with any set in $\cF$. A
data structure with the same name was later defined in \cite{mcc04} as
part of a {\em substitution decomposition} for a set system
$\cF$. \tnote{define substitutio decomp.} This is later used to define
a certificate if $\cF$ has no \COP. \PQRtree is a data structure
defined in \cite{mm96} as a data structure to represent any set system
$\cF$ with additional information in their $R$-nodes if $\cF$ has no
\COP.  All three of these data structures are proposed as
generalizations of \cite{bl76}'s \PQtree and hence produce the
\PQtree if $\cF$ has \COP. As a whole, all these three data
strutures are near-identical with their differences being notional.
In this section, we will discuss the basic theory that they all hold
and stick to the name of the data structure as \PQRtree.

A \PQRtree of a given set system $\cF$ from universe $U$ represents
all the sets from $\power{U}$ that have trivial intersections with
every set in $\cF$.  Two sets $A$ and $B$ are said to have a {\em
  trivial intersection} if $A \cap B$ is one of the following --
$\emptyset$, $A$ or $B$. In other words, $A$ and $B$ are either (i)
disjoint or (ii) one is the subset of the other. Such sets are also
called {\em orthogonal sets} or said that $A$ is orthogonal to
$B$. {\em Trivial subsets} of a universe $U$, denoted by $\cT(U)$, are
sets that have trivial intersections with any set in
$\power{U}$. Namely, $\cT(U)$ consists of $U$, singleton sets in
$\power{U}$ and $\emptyset$\cite{n89, mm96}.  In other words, the all
sets presented in a \PQRtree share the property that they do not
overlap with any set in $\cF$. We will now see how this helps in
testing for \COP in $\cF$.

An important observation made by \cite{mm96} is that if $\cF$ is a set
system with \COP then, aside from every set in $\cF$ being consecutive
(after applying the \COP order) the following sets must also be
consecutive for any $A, B \in \cF$.
\begin{enumerate}
  \item The intersection $A \cap B$ 
  \item The union $A \cup B$ if $A \cap B \ne \emptyset$
  \item The relative complements $A \setminus B$ and $B \setminus A$
    if $B \nsubseteq A$ and $A \nsubseteq B$ respectively.
\end{enumerate}
Also note that trivially, sets in $\cT(U)$ are consecutive in any
permutation of $U$\footnote{$\emptyset$ is considered consecutive by
  convention.}.

Any set system that is closed under the above operations and contains
$\cT(U)$ is called {\em complete} \cite{mm96} or a {\em weakly partative
  family} \cite{mcc04}. Such a set system is easier to
work with for reasons that will become apparent.\tnote{VAGUe!}
For any set system $\cF$, the smallest super set system that is
complete/weakly partative is called the {\em weak closure} of $\cF$,
denoted $\cW(\cF)$ \cite[Def.~3.2]{mcc04}. \cite[Def.~2, smallest {\em
  complete} supercollection]{mm96}.

What is interesting is that the set of \COP permutations of $\cF$ is
the same as that of $\cW(\cF)$ \cite[Th.~3]{mm96}. The reason this is
interesting is because it becomes apparent by this theory that it is
easier to use $\cW(\cF)$ to create the $\PQRtree$ for $\cF$.

The generalized \PQtree of $\cF$ is defined as the {\em decomposition
  tree} of its weak closure $\cW(\cF)$. Decomposition tree is the
inclusion tree (inclusion p.o. Hasse diagram) of the strong elements
of $\cW(\cF)$ with labels {\em prime, linear} and {\em degenerate}.\cite{mcc04}

Overlap components are computable in linear time \cite{mm95,wlh92}.

>>>>

\begin{definition}[{\em \Cop for set systems}]%
  \label{def:matrixsetsystem}%
  % [\emph{Set system of a binary matrix}.]
  Let $M$ be a binary matrix of order $n \times m$ and $\{c_i \mid i
  \in [m]\}$ be the columns in $M$.  A set system $\cF_{M} = \{S_i
  \mid S_i \subseteq [n], i \in [m]\}$ is defined such that for every
  column $c_i$ of $M$, set $S_i = \{j \mid m_{ji} = \un \}$. The
  collection $\cF_M$ is the {\em set system of binary matrix} $M$. The
  {\em binary matrix for set system} $\cF$ is conversely constructed
  and denoted by $M^\cF$. Thus, $M^{\cF_M} = M$.

  A set system $\cF$ from universe $U$, $|U| = n$ has the
  {\em \cop} if there exists a linear order or 
  permutation $\sigma = w_1w_2\ldots w_n$ that can be applied to $U$ \stt each set $S \in
  \cF$ becomes a consecutive subsequence\footnote{Also termed an {\em interval}} $w_{i}w_{i+1}\ldots
  w_{i+k-1}$ on $\sigma$ for some positive integer $i \le n+1-k$ where  $k = |S|$.

  The set $\valid{U,\cF}$ respresents all \COP orders of $\cF$ in $U$.
  \defnstop
\end{definition}

@@

Before proceeding to describe \PQRtree per se, we will see a few more
terminologies that will make the subsequent discussion in this section
simpler.

\begin{definition}[\em Orthogonal sets \cite{n89, mm96,
    mcc04}\footnote{\cite{mcc04} does not use the term ``mutually
    orthogonal'' but refers to the same idea as ``sets that do not
    overlap''. This terminology is also used in other literature like
    \cite{nsnrs09,wlh02}.} ]
  \label{def:orthosets}%
  Let $\cF$ be a set system with universe $U$ and sets $A, B \in \cF$.
  \begin{enumerate}
  \item $A$ and $B$ are said to have a {\em trivial intersection} if
    $A \cap B$ is one of the following.
    \begin{enumerate}[i.]
    \singlespacing
    \item $\emptyset$
    \item $A$
    \item $B$\footnote{In other words, $A$ and $B$ are either disjoint
        or one is the subset of the other.}
    \end{enumerate}
    % \vspace{-5mm}%

  \item $A$ and $B$ are called {\em mutually orthogonal} or $A$ is {\em
      orthogonal to} $B$ and vice versa, if they have a trivial
    intersection.

  \item {\em Trivial subsets} of a universe $U$, denoted by $\cT(U)$,
    are sets that have trivial intersections with any set in
    $\power{U}$. These sets are $U$, singleton sets in
    $\power{U}$ and $\emptyset$ \cite{n89, mm96}.
    Thus, $\cT(U) = \{U\} \  \bigcup \  \{ \{v\} \mid v \in U \} \  \bigcup \
    \{\emptyset\}$.

  \item {\em Orthogonal sets of a set system} $\cF$ with universe $U$
    are subsets of $U$ that are orthogonal to all sets in $\cF$. The
    set of all orthogonal sets to $\cF$ is denoted by
    $\cF^{\bot}$\footnote{\cite[Def.~3.1]{mcc04} uses the term {\em non-overlapping
    family} of $\cF$ and denotes it by $\cN(\cF)$. \label{mcc1}}.

  \item $\cF$ is called {\em complete}\footnote{\cite{mcc04} calls
      this a {\em weakly partative family}. \label{mcc2}} if the following hold
    true for every pair of non-orthogonal\footnote{Or overlapping.} sets
    $A, B$ in $\cF$.
    \begin{enumerate}[i.]
    \item $\cT(U) \subset \cF$
    \item $A \cup B \in \cF$
    \item $A \cap B \in \cF$
    \item $A \setminus B \in \cF$
    \item $B \setminus A \in \cF$
    \end{enumerate}
    In other words, $\cF$ contains all the trivial
    subsets of $U$, $A \cup B$ and the partitions of $A \cup B$
    defined by intersection and set difference.

  \item $\overline{\cF}$ represents the smallest super set system of
    $\cF$ that is complete\footnote{\cite[Def.~3.2]{mcc04} calls
      this the {\em weak closure} of $\cF$ denoted by $\cW(\cF)$. \label{mcc3}}

  \end{enumerate}
  \defnstop
\end{definition}

Generalized \PQtree or \gPQtree is a data structure defined in
\cite{n89} to represent all orthogonal sets of a set system $\cF$. A
data structure with the same name was later defined in \cite{mcc04} as
part of a {\em substitution decomposition} for a set system $\cF$
\tnote{define substitutio decomp.} and subsequently \cite{mcc04} gives
a new characterization of $\cF$ using a so-called incompatibility
graph. \PQRtree is defined in \cite{mm96} as a data
structure to represent any set system $\cF$ with additional
information in their $R$-nodes if $\cF$ has no \COP.  All three of
these data structures are proposed as generalizations of \cite{bl76}'s
\PQtree and hence produce the \PQtree if $\cF$ has \COP. As a whole,
all these three data strutures are largely identical with their
differences being notional.  In this section, we will discuss the
basic theory that they all hold. We will predominantly use the
terminology from \cite{mm96} and refer to the data structure as
\PQRtree.

An important observation made by \cite{mm96} is presented now along
with a few theorems that help in decomposing the \COP problem on $\cF$
into subproblems.

\begin{observation}[{\cite[Sec.~3]{mm96}}]
  If $\cF$ is a set system with \COP then, after applying the \COP
  order, not only must every set in $\cF$ be consecutive but the
  following sets must also be consecutive for any $A, B \in \cF$.
\begin{enumerate}
  \item The intersection $A \cap B$ 
  \item The union $A \cup B$ if $A \cap B \ne \emptyset$
  \item The relative complements $A \setminus B$ and $B \setminus A$
    if $B \nsubseteq A$ and $A \nsubseteq B$ respectively.
  \item Also note that trivially, sets in $\cT(U)$ are consecutive in
    any permutation of $U$\footnote{$\emptyset$ is considered
      consecutive by convention.}.
\end{enumerate}
\end{observation}

\begin{theorem}[{\cite[Th.~3,6]{mm96}}]
  \label{th:validcop}
  For any set system $\cF$ we have the following. \par
  \centering
    $\valid{\cF} = \valid{\overline{\cF}}$ \\
    $\cF^{\bot} = \overline{\cF^{\bot}} = (\overline{\cF})^{\bot}$
\end{theorem}

\begin{proposition}
  \label{pr:validcop}
  Any set that is consecutive on all the \COP permutations of $\cF$ is
  present in $\overline{\cF}$.
\end{proposition}

Proposition~\ref{pr:validcop} is owing to the following theorem by
which \cite{mm96} describes a way to decompose the problem of finding
all \COP orders of $\cF$ into two subproblems using sets in
$\overline{\cF} \cap \cF^{\bot}$.

\begin{theorem}[{\cite[Th.~7]{mm96}}]
  \label{th:mmdecomp}
  For any set system $\cF$, and $\emptyset \ne H \in \overline{\cF}
  \cap \cF^{\bot}$ we have the following.\par
  \centering
  $\valid{U,\cF} = \valid{U/H,\cF/H} * \valid{H,\cF \cap 2^H}$
\end{theorem}

\def \ff {\overline{\cF} \cap \cF^{\bot}}

The idea behind Theorem~\ref{th:mmdecomp} is as follows. A permutation $\alpha$
of $U$ is a composition of two permutations with respect to $H$ - i) a
permutation $\gamma$ of $H$ and (ii) a permutation $\beta$ of $U/H$.

For two mutually orthogonal sets $A, B$ such that $A \nsubseteq B$,
$A/B$ is defined as the set obtained by removing all elements of $B$
from $A$ and adding a representative element for $B$ in $A$. Being orthogonal,
results in only the following three possibilities.
\begin{enumerate}
\item $A$ and $ B$ are disjoint: $A/B  = A$
\item $A$ is a subset of $B$: $A/B$ is not defined
\item $B$ is a subset of $A$: $A/B = A \setminus B \cup \{b\}$, where $b$
is a new element not in $U$ added to represent $B$.
\end{enumerate}

We observe that this idea of decomposing the \COP problem into two
subproblems in \cite{mm96} is very similar to the substitution
decomposition of a set system given in \cite{mcc04}.\footnote{Using
  the theory cited in footnotes~\ref{mcc1},\ref{mcc2},\ref{mcc3}.}

The following corollary states how \PQRtree elegantly fits into this
whole theory and help in computing all \COP orders of $\cF$.

\begin{corollary}[{\cite[Cor.~8]{mm96}}]
  Let $\cF$ be a set system with universe $U$ and $H$ is a non-empty
  orthogonal set $H \in
  \ff$. If there is a \PQRtree $T_1$ that encodes all permutations in
  $\valid{U/H, \cF/H}$ and also a \PQRtree $T_2'$ that encodes all
  permutations $\valid{H,\cF \cap \power{H}}$, then a \PQRtree $T$ for
  $\cF$ can be obtained by replacing the leaf $h$ in $T_1$ by $T_2$.
\end{corollary}

%Finding elements in $\ff$ is done using \PQRtree.

Thus we have a recursive algorithm that can compute the \PQRtree for
$\cF$ provided we find an element from $\ff$ in each iteration. The
non-empty sets in $\ff$ are called {\em node sets} since they form the
nodes in the \PQRtree. They are calculated as follows.  One is by
computing the overlap components of $\cF$. The overlap components is
the partition that results from the {overlap} equivalence relation
which is nothing but non-orthogonal equivalence relation\footnote{It
  is easy to verify that this relation is indeed an equivalence
  relation.}. Overlap components are linearly
computable\cite{mm95,wlh92}. Once these elements are factored
out, the rest of the node sets are obtained by identifying {\em twin}
elements\footnote{\cite[Sec.~3]{kklv10} calls this {indistinguishable}
  elements. The equivalence class is called a {\em slot}.}. Two
elements $a, b \in U$ are twins if their membership in every set of
$\cF$ is in tandem with each other, \ie $\{a,b\} \bot \cF$. This is
clearly an equivalence relation and their equivalence classes is known
to be computable in linear time\cite{wlh01,mm96-ref11} and even in
logspace\cite{kklv10}.

The recursion end condition is when one cannot find any more sets from
$\ff$ that are non-trivial. This is when $\ff = \cT(U)$. This is the
point where the parents of the leaves of the \PQRtree are created. The
following theorem helps the algorithm decide whether a $P$, $Q$ or
$R$ node must be created.

\begin{theorem}[{\cite[Th.~9]{mm96},\cite[Th.~2.1]{mcc04}}]
\label{th:pqrfinalmm}
If $\cF$ is a set system with universe $U$ and $|U| \ge 3$ then one of
the following statements hold.
\begin{enumerate}
\item \label{th::it1} $\overline{\cF} = \cT(U)$ 
\item \label{th::it2}$\overline{\cF} = consec(\alpha)$\tnote{what is consec?} for
  some permutation $\alpha$ on $U$ 
\item \label{th::it3}$\overline{\cF} = \power{U}$ 
\end{enumerate}
\end{theorem}

In case (\ref{th::it1}) all elements in $U$ are made children of a
$P$ node. In case (\ref{th::it2}) all elements in $U$ are made
children of a $Q$ node in the order given by $\alpha$. Finally case
(\ref{th::it3}) is the one when no permutation of $U$ gives \COP. All
elements of $U$ are in this case made children in an $R$ node.

Theorem~\ref{th:pqrfinalmm} and the theory leading to it is very
similar to \cite[Th.~2.1, 3.5. Also Th.~3.2, 3.3, 3.4]{mcc04} which categorizes
the nodes in the above the three cases as {\em prime}, {\em linear}
and {\em degenerate} respectively. Their generalized \PQtree is
created in similar ways as \PQRtree above. This tree is in essense the
Hasse diagram of what they call {\em strong elements} of
$\overline{\cF}$. Strong elements of a set family are elements that do
not overlap with any other elements in the family, \ie it is
orthogonal to all other sets\cite[Def.~3.3]{mcc04}.

\begin{theorem}[{\cite{mm96}, \cite[Th.~3.6]{mcc04}}]
The set system $\cF$ has \COP \iff its \PQRtree has no $R$ nodes.  
\end{theorem}

% The generalized \PQtree of $\cF$ is defined as the {\em decomposition
%   tree} of its weak closure $\cW(\cF)$. Decomposition tree is the
% inclusion tree (inclusion p.o. Hasse diagram) of the strong elements
% of $\cW(\cF)$ with labels {\em prime, linear} and {\em degenerate}.\cite{mcc04}

% Overlap components are computable in linear time \cite{mm95,wlh92}.

----- Cert algo

------ ICPIA
